DATA 621---Business Analytics and Data Mining

### 2020-09-09 HW1/HW1_AF.Rmd
data exploration - initial

---
title: 'DATA 621 – Business Analytics and Data Mining: Homework #1'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg, Bryan Persaud,
  Jeff Shamp
date: "9/27/2020"
output:
  pdf_document: default
  html_document: default
---

@@

## 1. DATA EXPLORATION (Amber)
Describe the size and the variables in the moneyball training data set. Consider that too much detail will cause a manager to lose interest while too little detail will make the manager consider that you aren’t doing your job. Some suggestions are given below. Please do NOT treat this as a check list of things to do to complete the assignment. You should have your own thoughts on what to tell the boss. These are just ideas.

@@

@@

For the multiple linear regression model, will you use a metric such as Adjusted R2, RMSE, etc.? Be sure to explain how you can make inferences from the model, discuss multi-collinearity issues (if any), and discuss other relevant model output. Using the training data set, evaluate the multiple linear regression model based on (a) mean squared error, (b) R2, (c) F-statistic, and (d) residual plots. Make predictions using the evaluation data set.

@@

```{r}

train_set <- read.csv('data/moneyball-training-data.csv')
eval_set <- read.csv('data/moneyball-evaluation-data.csv')

colnames(train_set)

@@

>>>>

---
title: 'DATA 621---Business Analytics and Data Mining'
subtitle: 'Fall 2020---Group 2---Homework #1'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "9/27/2020"
output:
  pdf_document: default
  html_document: default
---

@@

```{r loadData, include=FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
train_set <- read.csv('data/moneyball-training-data.csv')
eval_set <- read.csv('data/moneyball-evaluation-data.csv')
```

## 1. DATA EXPLORATION (Amber)
Describe the size and the variables in the moneyball training data set. Consider
that too much detail will cause a manager to lose interest while too little
detail will make the manager consider that you aren't doing your job. Some
suggestions are given below. Please do NOT treat this as a check list of things
to do to complete the assignment. You should have your own thoughts on what to
tell the boss. These are just ideas.

@@

We are given a data set containing 2,276 records, each of which includes the 
number of wins the team had that year and 15 additional statistics about the 
season. A first look at the data shows that only about 8% of the records have 
a full set of data. The good news is that most of the missing values come from 
what we would expect; hit-by-pitch, caught stealing, double plays.

```{r, echo = FALSE, message = FALSE, warning = FALSE}

pct_null <- data.frame(do.call("rbind", map(train_set %>% select(-INDEX), ~mean(is.na(.)))))
colnames(pct_null) <- c('PCT_NULL')

pct_null %>%
  arrange(desc(PCT_NULL)) %>%
  filter(PCT_NULL > 0)

  * Total cases: 2276, Complete cases: 191 (only ~8% of the data has fully filled in values)
  * 15 stats (variables), 1 target (wins)
  * Best team won 146 out of 162 games, worst team won none, Average number of wins is 80 (show win dist)
  * Discuss missing values

```{r, echo = FALSE, message = FALSE, warning = FALSE}

# histogram of wins
p <- ggplot(train_set, aes(x=TARGET_WINS)) + 
  geom_histogram() + 
  geom_vline(aes(xintercept=mean(TARGET_WINS)),
            color="blue", linetype="dashed", size=1)+
  labs(title="Distribution of Wins",x="Number of wins", y = "Number of teams")

p

# Correlation matrix with significance levels (p-value)
library("Hmisc")
res2 <- rcorr(as.matrix(train_set %>% select(-INDEX)))

# Insignificant correlation are crossed
corrplot(res2$r, type="upper", order="hclust", 
          tl.col = "black", p.mat = res2$P, sig.level = 0.01, insig = "blank")

train_set %>%
  select(-INDEX) %>%
  gather(-TARGET_WINS, key = "STAT", value = "VALUE") %>%
  ggplot(aes(x = VALUE, y = TARGET_WINS)) +
    geom_point() +
  labs(title="Number of wins with respect to recorded stats",x="", y = "Number of Wins") +
    facet_wrap(~ STAT, scales = "free")

@@

For the multiple linear regression model, will you use a metric such as Adjusted
\(R^2\), RMSE, etc.? Be sure to explain how you can make inferences from the
model, discuss multi-collinearity issues (if any), and discuss other relevant
model output. Using the training data set, evaluate the multiple linear
regression model based on (a) mean squared error, (b) \(R^2\), (c) F-statistic,
and (d) residual plots. Make predictions using the evaluation data set.

@@

```{r initial, warning = FALSE}
########### libraries 
library(tidyverse)

########### data
train_set <- read.csv('data/moneyball-training-data.csv')
eval_set <- read.csv('data/moneyball-evaluation-data.csv')

@@

```{r dataExploration, results = 'hide'}
# initial summary of all values 
summary(train_set)
p <- ggplot(train_set, aes(x=TARGET_WINS)) + 
  geom_histogram() + 
  geom_vline(aes(xintercept=mean(TARGET_WINS)),
            color="blue", linetype="dashed", size=1)+
  labs(title="Distribution of Wins",x="Number of wins", y = "Number of teams")

# complete vs total cases
train_set %>%
  filter(complete.cases(.) == TRUE) %>%
  count() %>%
  rename("complete cases"= n) %>%
  bind_cols(
    train_set %>%
    select(TARGET_WINS) %>%
    count() %>%
    rename("target count" = n)
  )
```
